Week 4 Assignment - Using Movies data set
Which year were the best popular movies made?

========================================================
Definitions:
'Best' movies are defined here as movies with the highest average rating within each genre. It is important to define what is best for each genre because each genre has a distinct mean, and a 'best' average rating in one particular genre may not be a 'best' rating in another.

'Popular' movies are defined here as movies receiving a high number of votes, inferring that movies with a large number of votes were seen by a corresponding large number of people, and can be considered popular. 
Both the average rating and the number of votes are normally distributed. 


### Import Data
```{r set-options, echo=FALSE, cache=FALSE}
options(width=120)

library(dplyr)
library(ggplot2)
library(reshape2)

(df.movies <- tbl_df(read.table("movies.tab", sep="\t", header=TRUE, quote="", comment="")))

```
### Each genre has a different distribution of ratings and votes, so it is important to consider them seperatly

Votes - comparison of means
The anova results in rejection of the hypothesis that the theoretical mean is the same across genres
A simple bar chart illustrates the variability between genres
```{r}
genres = seq(length(df.movies)-6,length(df.movies))
names(genres) = tail(colnames(df.movies), 7)

m.vote <- filter(group_by(melt(select(df.movies,votes, rating, genres),
                               id =  c("rating", "votes")),variable),value!=0)
m.vote  <- filter(melt(select(df.movies, votes, genres), id = "votes"), value != 0)
aov.out <- aov(value ~ variable, data = m.vote)
summary.lm(aov.out)
s.len <- summarise(group_by(m.vote, variable), avg.vote = mean(votes))
c <- ggplot(data = s.len, aes(x = variable, y = avg.vote, fill = variable)) 
c + geom_bar(stat = "identity") + geom_hline(aes(yintercept = mean(s.len$avg.vote)),colour = "red")

```

Ratings - comparison of means
The anova results in rejection of the hypothesis that the theoretical mean is the same across genres
A simple bar chart illustrates the variability between genres
```{r}
m.rate  <- filter(melt(select(df.movies, rating, genres), id = "rating"), value != 0)
aov.out <- aov(value ~ variable, data = m.rate)
summary.lm(aov.out)
s.len <- summarise(group_by(m.rate, variable), avg.rate = mean(rating))
ggplot(data = s.len, aes(x = variable, y = avg.rate, fill = variable)) + geom_bar(stat = "identity" ) + geom_hline(yintercept = mean(s.len$avg.rate), colour = "red")
                                                                                                            
```

### Selecting and plotting a subset of data to include just those movies that meet the folowing criteria:

Number of votes is in the upper 1%  for the genre
Rating is in the upper 1% for the genre

# Calculate the cutoff for the upper 95% of votes for each genre
```{r norm_figs, fig.width = 8, fig.height = 6, dev='svg'}

#Dist. of votes, only intersted in uppper tail
summary(df.movies$votes)

m.vote <- filter(group_by(melt(select(df.movies,votes, genres), id =  "votes"), 
                       variable),value!=0)

# looking at top 95% of the votes

print("Table of 95 percentile of votes by genre")

(vote.95 <- summarise(m.vote, cutoff.v = quantile(votes, .95)))
cuts.v <- vote.95$cutoff.v
names(cuts.v) = vote.95$variable


```

### 95% percentile rating for each genre
```{r cutoffrate}
#summary of all rating, only interested in upper tail
summary(df.movies$rating)

m.rate <- filter(group_by(melt(select(df.movies, rating, genres), id =  "rating"), 
                          variable),value!=0)
print("Table of 95 percentile of ratings by genre")
(rate.95 <- summarise(m.rate, cutoff.r = quantile(rating, .95)))
cuts <- rate.95$cutoff
names(cuts) = rate.95$variable
```

Select the 'best' movies in each genre

```{r bests}
# year + title identifies the movie
m <- select(df.movies, title, year, genres, votes, rating)
mtop <- filter(melt(m, id =  c("rating", "votes", "year", "title")), value != 0)
mtop  <- merge(mtop, rate.95,  by.mtop = "variable", by.rate.95 = "variable")
mtop  <- filter(merge(mtop, vote.95,  by.mtop = "variable", by.vote.95 = "variable"), rating >= cutoff.r, votes >= cutoff.v)

# organizaing DAta

best.movies <- cbind(select(mtop, variable, title, year, rating, votes), score = mtop$rating * mtop$votes/1000, titleyear = paste(mtop$title, " - ", mtop$year))
colnames(best.movies)[1] = 'genre'
xx <- best.movies[order(best.movies$titleyear, -best.movies$score),]
xx <- xx[!duplicated(xx$titleyear),]

```
### The list of best popular movies, sorted by genre and then score
There is no distinct film in the genre Documentary that meets the criteria of being in the 95th percentile for both rating and votes
Listing the top five in each
```{r}

list.of.movies <- select(xx[order(xx$genre, -xx$score),], genre, title, year, rating, votes, score)

# top five in each
(by(data = list.of.movies, INDICES = list.of.movies$genre, FUN = function(x) head(x, 5)))

print("Count by genre, Documentary has no distinct film that meets both criteria")

print(table(list.of.movies$genre))


print("Summary Stats")
print(summarise(group_by(list.of.movies, genre), total.titles = length(titleyear), max.rate = max(rating), max.votes= max(votes), max.total = max(score)))

```

